---
title: "Overrepresentation"
author: "Dan MacKinnon"
date: "February 14, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Overrepresentation occurs when a group within a population is disproportinately found within a subset of the population. When applied to human groups, overrepresenation is often indicative of an underlying bias or social issue. For example:

> Indigenous people make up 4% of the Canadian population. However, 20% of female victims of homicide in Canada are Indigenous.

> Though African Americans and Hispanics make up approximately 32% of the US population, they comprised 56% of all incarcerated people in the US.

> Men represent about 49% of the US population, but represent 72.2% of appointments to boards of directors of Fortune 500 companies.

In each case, a particular group is represented at a different level within a particular subset of a population than it is represented in the general population 

When we make statements about overrepresentation, sometimes we want to say tings like "A man is X times more likely than a woman to be appointed to a board of directors." However, it is not immediately obvious how to make such a statement based on the data provided. In this notebook, we'll look at how we can use conditional probability to make such statements.

## Problem statement

>There is a group X that makes up 4% of a population. Within the population there is a subset Y, and group X makes up 20% of that subset. We would say that X is overrepresented in the subset Y. How much more likely is it that a person of group X will end up in subset Y compared to someone not in group X?

## Direct Calculation

The problem can be expressed in terms of conditional probabilities. The probability of an event A given that an event B has already occured is called the conditional probablility of A given B, is written as $p(A|B)$ and has the formula:

$$p(A|B) = \frac{p(A \cap B)}{p(B)}$$



If we say that $X$ is the event that a person is in group X, and $Y$ is the event that a person is in subset Y, we know the probability of being in group X is $p(X) = 0.04$, and the probability of being in group X given that they are in subset Y, $p(X|Y) = 0.8$. Also, we know the corresponding probabilities for those not in group X, $p(\neg X) = 0.96$ and $p(\neg X|Y) = 0.2$.

What we would like to calculate is the probability of someone being in subset Y given that they are in group X, compared to the probability of being in subset Y given that they are not in group X. We can express this as:

$$\frac{p(Y|X)}{p(Y|\neg X)}$$ 
First, we can find a formula that expresses how being a member in group X affects the probabilty of being in subset Y. 

$$p(X|Y) = \frac{p(X \cap Y)}{p(Y)} = 0.2$$
So, $p(X \cap Y) = 0.2 \times p(y)$. To obtain an expression for $p(Y|X)$, we start with the conditional probability formula, and substitute in the value we know for $p(X)$.
$$p(Y|X) = \frac{p(X \cap Y)}{p(X)} = \frac{p(X \cap Y)}{0.04}$$
$$p(Y|X) = \frac{p(X \cap Y)}{0.04} = \frac{0.2p(Y)}{0.04} = 5p(Y) $$
This formula is expressing how being a member of the group X impacts the probability of being in subset Y. In this particular case, being in group X makes it 5 times more likely to be in subset Y, when compared to the 'raw' probability of being in subset Y.

The same sort of calculation can tell us how not being a member of the group X impacts the probability of being in subset Y.


$$p(\neg X|Y) = \frac{p(Y \cap \neg X)}{p(Y)} = 0.8$$

So, $p(Y \cap \neg X) = 0.8p(Y)$, and substituting this into the formula for $p(Y| \neg X)$ we obtain:

$$p(Y|\neg X) = \frac{p(Y \cap \neg X)}{p(\neg X)} =  \frac{p(Y \cap \neg X)}{0.96} =  \frac{0.8p(Y)}{0.96} = \frac{5}{6}p(Y)$$
Combining these, 

$$\frac{p(Y|X)}{p(Y|\neg X)} = \frac{5p(Y)}{\frac{5}{6}p(Y)} = 6$$

Based on these calculations, we would say that a person belonging to group X is 6 times more likely than someone not in group X to belong to subset Y.


## Simulation

Returning to the problem statement:

>There is a group X that makes up 4% of a population. Within the population there is a subset Y, and group X makes up 20% of that subset. We would say that X is overrepresented in the subset Y. How much more likely is it that a person of group X will end up in subset Y compared to someone not in group X?

```{r}
base_rate <- 0.1
x_rate <- 5*base_rate
nx_rate <- 0.8*base_rate
trial <- 1:5000
OverRep <- data.frame(trial)
OverRep$memberX <- (runif(5000,0,1) <= 0.04)
for (i in trial){
  if (OverRep$memberX[i]){
    OverRep$memberY[i] <- runif(1,0,1) <= x_rate
  } else {
    OverRep$memberY[i] <- runif(1,0,1) <= nx_rate
  }
}

```
As designed, about 4% is in group X:
```{r}
prop.table(table(OverRep$memberX))
```
We should see that the base rate of membership in subset Y for the whole population is about 10%.
```{r}
prop.table(table(OverRep$memberY))
```
However, members of group X make up a disproportionate amount of the subset Y.
```{r}
t <- table(OverRep$memberX, OverRep$memberY)
barplot(t, main="Representation in group X and subset Y",
  xlab="", col=c("light grey","dark grey"),
 	legend = c('in X', 'in Y'))
```

About 20% of subset Y should be members of the group X.
```{r}
InSubsetY <- subset(OverRep, memberY==TRUE)
prop.table(table(InSubsetY$memberX))
```

## General formula

The quotient described above  general measure of the overrepresentation of the group X in subset Y:

$$OverRep(X,Y) = \frac{p(Y|X)}{p(Y|\neg X)} $$

Using the numbers usually presented when talking about overrepresentation if $p$ is the fraction of the population in group X, and $q$ is the fraction of the population in subset Y that is in group X, we have:

$$OverRep(X,Y) = \frac{q}{p}\times\frac{1-p}{1-q} $$
In the example above this is
$$OverRep(X,Y) = \frac{0.2}{0.04}\times\frac{0.96}{0.8} = 6 $$

Like most statistics, this can be misleading when used to describe individuals. Consider the problem of the male and female CEOs mentioned above:

> Men represent about 49% of the US population, but represent 72.2% of appointments to boards of directors of Fortune 500 companies.

If $M$ is being male and $C$ is bieng a CEO:

$OverRep(M,C) = \frad{0.722}{0.49}\times \frac{0.51}{0.278} = 2.7$

If we consider Joe and Jane, where Joe is a 40 year old high school drop out, and Jane is a 40 year old MBA graduate with a law degree, is Joe almost 3 times more likely to be a CEO than Jane? No. But what the "over representation quotient" does say is that there is a discrepancy between male and female representation among CEOs, and the 2.7 provides a measure of how acute that discrepancy is.
