---
title: "Deep Dive into the Multiplication Table, with R"
author: "dmackinnon1"
date: "November 26, 2015"
output: html_document
---

## Generating a 10x10 multiplication table in R
Just simple explorations of the multiplication table in R, originally shared at [mathrecreation.com](http://www.mathrecreation.com/2012/08/a-deep-dive-into-multiplication-table.html) using Fathom and Tinkerplots.

There are many ways of generating a 10x10 multiplicaiton table, here's one:

```{r}
mult <- matrix(1:10, 1,10)
for (i in 2:10) {
  mult <- rbind(mult, i*mult[1,])
}
mult
```

## Multiplication Rainbow, in R
And here's the rainbow.

```{r, echo=FALSE}
n <- 100
x <- seq(1:10) 
y <- x
rb <- rainbow(n, s = 1, v = 1, start = 0, end = max(1, n - 1)/n, alpha = 1)
image(x=x, y=y, mult, col=rb, main="Multiplication Rainbow", ylim = rev(range(y)))
```

## Star Polygons in the multiplication table

This one is a little convoluted in R.
```{r}
skipCount <- append(mult[3,],3) %%10
circleY <- sin(2*pi*skipCount/10)
circleX <- cos(2*pi*skipCount/10)
plot(circleX, circleY, type = "l", main="Star Polygon from Skip Counting", axes=FALSE, ylab = "", xlab = "")
```

## Modular patterns in the multiplication table
```{r}
mod <- mult %%5
mod
image(x=x, y=y, mod, col=rb, main="Modular Multiplication Quilt", ylim = rev(range(y)))
```